---
title: "NOAA Storms Analysis"
author: "ACBLimehouse"
date: "6/18/2017"
output: html_document
---
## Analysis Generated for Coursera Course: Reproducible Research  
  
### Section One: Setup and Data Import  
  
```{r setup, include=FALSE, root.dir = "/Users/adamlimehouse/Desktop/Dropbox/03 Projects Folder/Economic and Policy Analysis/Intro to Data Science R Files/Reproducible Research - Peer Graded Assessments/RR-WK4-CP-2"}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy = TRUE)
MyWD <- "/Users/adamlimehouse/Desktop/Dropbox/03 Projects Folder/Economic and Policy Analysis/Intro to Data Science R Files/Reproducible Research - Peer Graded Assessments/RR-WK4-CP-2"
setwd(MyWD)
```
  
We begin the analysis by downloading and importing the Storms Data.  
  
```{r initial import, results="hide", message=FALSE, warning=FALSE}
dataURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
documentationURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
documfile <- "StormDataDocumentation.PDF"
compfilename <- "StormData.csv.bz2"
filename <- "StormData.csv"
DataName <- "StormData"

library(R.utils)
library(dplyr)

## Download the Documentation for reference
if (!file.exists(documfile)){
  download.file(documentationURL, documfile)
}

## Download the Data and Read it into the Global Environment
if (!file.exists(filename)){
  if (!file.exists(compfilename)){
    download.file(dataURL, compfilename, method="curl")
  }
  bunzip2(compfilename, filename, remove = FALSE, skip = TRUE)
}
if (!exists(DataName)){
  StormData <- read.csv(filename, header = TRUE)
}
```
  
### Section Two: Data Review and Cleaning
  
After importing the data, we take a look at it to get a sense of how it needs to be cleaned in order to do a reasonable analysis. First, we look at the column names and top of the data set.
  
```{r Data Review 1}
names(StormData)
head(StormData)
```
  
Then, we take a look at the top 15 most frequent event types in NOAA's Storm Data.
  
``` {r Data Review 2}
event_types <- as.data.frame(table(StormData$EVTYPE))
event_types <- event_types[order(-event_types$Freq), ]
head(event_types, 15)
rm(event_types)
```
  
This tells us that there are many event type names that overlap in commonly understood categories. Some of these categories include:  
- Cold Weather;  
- Tornados;  
- Nautical events including high seas, heavy waves, etc;  
- Tropical Storms including typhoons, hurricanes, and tropical depressions;  
- Wind; and,  
- Flood  
  
So, with that in mind, we go about combining the event types using R's grepl. __Note__: in cases of event type categories that include "/", "&", or "and", those event types are collapsed into a single value named "Multiple Event". This "Multiple Event" category is included in the analysis for completeness but it should be understood that it overlaps most of the other singular event types.  
  
```{r Data Cleaning 1}
StormData$EVTYPE <- as.character(StormData$EVTYPE)
StormData$EVTYPE[grepl("/|&|and", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Multiple Event"
StormData$EVTYPE[grepl("volc", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Volcano"
StormData$EVTYPE[grepl("DROUGHT", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Drought"
StormData$EVTYPE[grepl("tsunami", StormData$EVTYPE,ignore.case = TRUE)] <-
  "Tsnumai"
StormData$EVTYPE[grepl("LIGHTNING", StormData$EVTYPE,ignore.case = TRUE)] <-
  "Lightning"
StormData$EVTYPE[grepl("Avalanc*", StormData$EVTYPE,ignore.case = TRUE)] <-     
  "Avalanche"
StormData$EVTYPE[grepl("*Fog", StormData$EVTYPE,ignore.case = TRUE)] <-
  "Fog"
StormData$EVTYPE[grepl("wind|wnd", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Wind"
StormData$EVTYPE[grepl("funnel|tornado", StormData$EVTYPE,ignore.case = TRUE)] <-
  "Tornado"
StormData$EVTYPE[grepl("glaze", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Glaze"
StormData$EVTYPE[grepl("hail", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Hail"
StormData$EVTYPE[grepl("dust", StormData$EVTYPE,ignore.case = TRUE)]  <- 
  "Dust"
StormData$EVTYPE[grepl("flood|*rising*|high water", StormData$EVTYPE,ignore.case = TRUE)] <- "Flood"
StormData$EVTYPE[grepl("ic(e|y)", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Ice"
StormData$EVTYPE[grepl("fire|smoke", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Fire"
StormData$EVTYPE[grepl("thunder", StormData$EVTYPE,ignore.case = TRUE)] <- "Non-Tropical Storm"
StormData$EVTYPE[grepl("Coastal*|Non-Tropical*|microburst", StormData$EVTYPE, ignore.case = TRUE)] <- 
  "Non-Tropical Storm"
StormData$EVTYPE[grepl("slide|eros", StormData$EVTYPE,ignore.case = TRUE)] <-
  "Erosion"
StormData$EVTYPE[grepl("rain", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Rain"
StormData$EVTYPE[grepl("freez|cold|snow|chill|winter|blizzard|sleet|glaze|frost|hypothermia|wintry*|low temp*|mixed precip|ice", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Cold Weather"
StormData$EVTYPE[grepl("TROPICAL.Storm|typhoon|*depression", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Tropical Storm"
StormData$EVTYPE[grepl("(hurri|opal)", StormData$EVTYPE,ignore.case = TRUE)] <- "Tropical Storm"
StormData$EVTYPE[grepl("RIP CURREN(T|TS)", StormData$EVTYPE,ignore.case = TRUE)] <-
  "Rip Current"
StormData$EVTYPE[grepl("*Current|*SURF|*SEAS|marine*|*swells|*waves|waterspout|*tide", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Nautical"
StormData$EVTYPE[grepl("heat", StormData$EVTYPE,ignore.case = TRUE)] <- 
  "Heat"
StormData$EVTYPE[grepl("Storm SURGE", StormData$EVTYPE,ignore.case = TRUE)] <-
  "Storm Surge"
```
  
Now we can take a look at the Property Damage and Crop Damage Exponents to get a sense of they are used in the data set. We start by examining and then transforming the Property Damage Exponents variable.  
  
``` {r Data Cleaning 2}
## Clean the Property and Crop Damage Data
unique(StormData$PROPDMGEXP)

## Transform the property damage exponents into numeric values
StormData$PROPDMG[StormData$PROPDMGEXP == "K"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "K"] * 1000
StormData$PROPDMG[StormData$PROPDMGEXP == "M"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "M"] * (10^6)
StormData$PROPDMG[StormData$PROPDMGEXP == "H"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "H"] * 100
StormData$PROPDMG[StormData$PROPDMGEXP == "h"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "h"] * 100
StormData$PROPDMG[StormData$PROPDMGEXP == ""] <- StormData$PROPDMG[StormData$PROPDMGEXP == ""] * 1
StormData$PROPDMG[StormData$PROPDMGEXP == "B"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "B"] * (10^9)
StormData$PROPDMG[StormData$PROPDMGEXP == "m"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "m"] * (10^6)
StormData$PROPDMG[StormData$PROPDMGEXP == "0"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "0"] * 1
StormData$PROPDMG[StormData$PROPDMGEXP == "1"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "1"] * 10
StormData$PROPDMG[StormData$PROPDMGEXP == "2"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "2"] * 100
StormData$PROPDMG[StormData$PROPDMGEXP == "3"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "3"] * 1000
StormData$PROPDMG[StormData$PROPDMGEXP == "4"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "4"] * (10^4)
StormData$PROPDMG[StormData$PROPDMGEXP == "5"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "5"] * (10^5)
StormData$PROPDMG[StormData$PROPDMGEXP == "6"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "6"] * (10^6)
StormData$PROPDMG[StormData$PROPDMGEXP == "7"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "7"] * (10^7)
StormData$PROPDMG[StormData$PROPDMGEXP == "8"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "8"] * (10^8)
StormData$PROPDMG[StormData$PROPDMGEXP == "+"] <- 0
StormData$PROPDMG[StormData$PROPDMGEXP == "-"] <- 0
StormData$PROPDMG[StormData$PROPDMGEXP == "?"] <- 0
head(StormData[,c("EVTYPE","PROPDMG","PROPDMGEXP")])
```
  
As you can see, NOAA's meterologists used a variety of means to indicate exponents including "K" and "3" for thousands, and "B" and "8" for billions. We can now perform the same tranformation for the Crop Damage Exponent.  
  
``` {r Data Cleaning 3}
## Crop Damage transformation
StormData$CROPDMG[StormData$CROPDMGEXP == "M"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "M"] * (10^6)
StormData$CROPDMG[StormData$CROPDMGEXP == "K"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "K"] * 1000
StormData$CROPDMG[StormData$CROPDMGEXP == "m"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "m"] * (10^6)
StormData$CROPDMG[StormData$CROPDMGEXP == "B"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "B"] * (10^9)
StormData$CROPDMG[StormData$CROPDMGEXP == "k"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "k"] * 1000
StormData$CROPDMG[StormData$CROPDMGEXP == "0"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "0"] * 1
StormData$CROPDMG[StormData$CROPDMGEXP == "2"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "2"] * 100
StormData$CROPDMG[StormData$CROPDMGEXP == ""] <- StormData$CROPDMG[StormData$CROPDMGEXP == ""] * 1
StormData$CROPDMG[StormData$CROPDMGEXP == "?"] <- 0
head(StormData[,c("EVTYPE","CROPDMG","CROPDMGEXP")])
```
  
We're now in a position to take a look at the data and answer two questions:  
1. __Across the US, which event typs (EVTYPE variable) are most harmful to population health?__  
2. __Across the US, which event types have the greatest economic conseqences?__  
  
## Section Three: Data Analysis and Results  
    
### Public Health Consequences  
  
Let's address the population health consequences first. Usually, when we think of public health we include chronic and acute disease, injury, and fatalities. Unfortunately, NOAA's data set does not include the disease consequences of crop damage, dust, flooding, etc. Instead, the public health consequences of meteorological events are reduced only to injuries and fatalities.  
  
So, we aggregate the top 10 meteorological causes of both.

```{r Analysis 1}
## Fatalities aggregation
fatal <- aggregate(FATALITIES~EVTYPE,data=StormData,FUN=sum,na.rm=TRUE)
fatal <- fatal[with(fatal,order(-FATALITIES)),]
fatal <- head(fatal, 10)
print(fatal)

## Injuries aggregation
injury <- aggregate(INJURIES~EVTYPE,data=StormData,FUN=sum,na.rm=TRUE)
injury <- injury[with(injury,order(-INJURIES)),]
injury <- head(injury,10)
print(injury)
```
  
As you can see, Tornados are the event type that causes the most fatalities and injuries across the whole of the US. Now, let's take a look at this graphically.  

``` {r Analysis 2}
##plot the graph showing the top 10 fatalities and injuries
par(mfrow=c(1,2),mar=c(8,5,4,2))
barplot(fatal$FATALITIES/(10^3),names.arg=fatal$EVTYPE,las=2,col="black",ylab="Fatalities (Thousands)",main="Top 10 Causes of Fatalities")
barplot(injury$INJURIES/(10^3),names.arg=injury$EVTYPE,las=2,col="red",ylab="Injuries (Thousands)",main="Top 10 Causes of Injuries")
```
  
Again, Tornados are far and away the most important cause of fatalities and injuries in the US. This makes some sense as, according to NOAA, more than 1,200 tornadoes strike the US every year. See [here.](http://www.nssl.noaa.gov/education/svrwx101/tornadoes/)  

### Economic Consequences
Now we can take a look at the economic consequences of the meteorological events in the NOAA data set. These consequences include property damage and crop damage. Why, you might ask, are these separated? I don't know. My guess is that the data for these are tracked by separate agencies of the US government (for crop data, it's probably the [USDA](https://usdasearch.usda.gov/search?utf8=%E2%9C%93&affiliate=usda&query=crop+damage+due+to+weather)).  
  
Regardless, let's perform the same aggregation as before on property and crop damage and see what happens.  
  
```{r Analysis 3}
## Economic Consequences
prop <- aggregate(PROPDMG~EVTYPE,data=StormData,FUN=sum,na.rm=TRUE)
prop <- prop[with(prop,order(-PROPDMG)),]
prop <- head(prop,10)
print(prop)
crop <- aggregate(CROPDMG~EVTYPE,data=StormData,FUN=sum,na.rm=TRUE)
crop <- crop[with(crop,order(-CROPDMG)),]
crop <- head(crop,10)
print(crop)
```
  
As you can see, Floods cause far and away more property damage than any other type of event while drought, which can last for years, causes more damage to crops. However, this may not be a fair analysis since, as I said, droughts take place over a long period of time, while floods, which cause almost as much damage, take place over a comparatively short time scale.  
  
To wrap up, let's show this graphically:  
  
```{r Analysis 4}
## Economic Results 
par(mfrow=c(1,2),mar=c(8,5,4,2))

##plot the graph showing the top 10 property and crop damages
barplot(prop$PROPDMG/(10^9),names.arg=prop$EVTYPE,las=2,col="green",ylab="Prop.damage (Billions)",main="Top 10 Causes of Prop. Damage")
barplot(crop$CROPDMG/(10^9),names.arg=crop$EVTYPE,las=2,col="yellow",ylab="Crop damage (Billions)",main="Top 10 Causes of Crop Damage")
```

_There you have it._ 

## Conclusions
To sum up this analysis, here are three conclusions and a caveat:  
1 Tornados are bad news and cause the most deaths and injuries of any kind of weather in the US.  
2 Floods, while they don't kill and maim as many people, are not great for property or crops.  
3 Finally, Drought also quite terrible for crops but may be less important than appears because of a time-scale issue.  
  
The caveat is this: The time-scale issue may not actually be an issue since we have measured the consequences of these events across a comparable time period. However, my sense is that due to the slow moving nature of draught it is easier to measure the damage done to crops from a baseline year-over-year average.  
  
Questions, Comments, Elucidations of measuring crop damage due to fire or flood? Email me [here](mailto:alimehouse@gmail.com) or check out the git hub page where you can find the R code and Rmd files I used to generate the analysis [here](https://github.com/acblimehouse/RR-WK4-CP-2).