---
title: "NOAA Storms Analysis"
author: "ACBLimehouse"
date: "6/18/2017"
output: html_document
---
## Analysis Generated for Coursera Course: Reproducible Research

### Section One: Setup and Data Import

```{r setup, include=FALSE, root.dir = "/Users/adamlimehouse/Desktop/Dropbox/03 Projects Folder/Economic and Policy Analysis/Intro to Data Science R Files/Reproducible Research - Peer Graded Assessments/RR-WK4-CP-2"}
knitr::opts_chunk$set(echo = TRUE)
MyWD <- "/Users/adamlimehouse/Desktop/Dropbox/03 Projects Folder/Economic and Policy Analysis/Intro to Data Science R Files/Reproducible Research - Peer Graded Assessments/RR-WK4-CP-2"
setwd(MyWD)
```

We begin the analysis by downloading and importing the Storms Data.

```{r initial import}
dataURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
documentationURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
documfile <- "StormDataDocumentation.PDF"
compfilename <- "StormData.csv.bz2"
filename <- "StormData.csv"
DataName <- "StormData"

library(R.utils)
library(dplyr)

## Download the Documentation for reference
if (!file.exists(documfile)){
  download.file(documentationURL, documfile)
}

## Download the Data and Read it into the Global Environment
if (!file.exists(filename)){
  if (!file.exists(compfilename)){
    download.file(dataURL, compfilename, method="curl")
  }
  bunzip2(compfilename, filename, remove = FALSE, skip = TRUE)
}
if (!exists(DataName)){
  StormData <- read.csv(filename, header = TRUE)
}
```

### Section Two: Data Review and Cleaning

After importing the data, we take a look at it to get a sense of how it needs to be cleaned in order to do a reasonable analysis. First, we look at the column names and top of the data set.

```{r Data Review 1}
names(StormData)
head(StormData)
```

Then, we take a look at the top 15 most frequent event types in NOAA's Storm Data.

``` {r Data Review 2}
event_types <- as.data.frame(table(StormData$EVTYPE))
event_types <- event_types[order(-event_types$Freq), ]
head(event_types, 15)
rm(event_types)
```

This tells us that there are many event type names that overlap in commonly understood categories. Some of these categories include:  
- Cold Weather;  
- Tornados;  
- Nautical events including high seas, heavy waves, etc;  
- Tropical Storms including typhoons, hurricanes, and tropical depressions;  
- Wind; and,  
- Flood  

So, with that in mind, we go about combining the event types using R's grepl. Note, in cases of event type categories that include "/", "&", or "and", those event types are collapsed into a single value named "Multiple Event". This "Multiple Event" category is included in the analysis for completeness but it should be understood that it overlaps most of the other singular event types.

```{r Data Cleaning 1}
StormData$EVTYPE <- as.character(StormData$EVTYPE)
StormData$EVTYPE[grepl("/|&|and", StormData$EVTYPE,ignore.case = TRUE)] <- "Multiple Event"
StormData$EVTYPE[grepl("volc", StormData$EVTYPE,ignore.case = TRUE)] <- "Volcano"
StormData$EVTYPE[grepl("DROUGHT", StormData$EVTYPE,ignore.case = TRUE)] <- "Drought"
StormData$EVTYPE[grepl("tsunami", StormData$EVTYPE,ignore.case = TRUE)] <- "Tsnumai"
StormData$EVTYPE[grepl("LIGHTNING", StormData$EVTYPE,ignore.case = TRUE)] <- "Lightning"
StormData$EVTYPE[grepl("Avalanc*", StormData$EVTYPE,ignore.case = TRUE)] <- "Avalanche"
StormData$EVTYPE[grepl("*Fog", StormData$EVTYPE,ignore.case = TRUE)] <- "Fog"
StormData$EVTYPE[grepl("wind|wnd", StormData$EVTYPE,ignore.case = TRUE)] <- "Wind"
StormData$EVTYPE[grepl("funnel|tornado", StormData$EVTYPE,ignore.case = TRUE)] <- "Tornado"
StormData$EVTYPE[grepl("glaze", StormData$EVTYPE,ignore.case = TRUE)] <- "Glaze"
StormData$EVTYPE[grepl("hail", StormData$EVTYPE,ignore.case = TRUE)] <- "Hail"
StormData$EVTYPE[grepl("dust", StormData$EVTYPE,ignore.case = TRUE)]  <- "Dust"
StormData$EVTYPE[grepl("flood|*rising*|high water", StormData$EVTYPE,ignore.case = TRUE)] <- "Flood"
StormData$EVTYPE[grepl("ic(e|y)", StormData$EVTYPE,ignore.case = TRUE)] <- "Ice"
StormData$EVTYPE[grepl("fire|smoke", StormData$EVTYPE,ignore.case = TRUE)] <- "Fire"
StormData$EVTYPE[grepl("thunder", StormData$EVTYPE,ignore.case = TRUE)] <- "Non-Tropical Storm"
StormData$EVTYPE[grepl("Coastal*|Non-Tropical*|microburst", StormData$EVTYPE,
                       ignore.case = TRUE)] <- "Non-Tropical Storm"
StormData$EVTYPE[grepl("slide|eros", StormData$EVTYPE,ignore.case = TRUE)] <- "Erosion"
StormData$EVTYPE[grepl("rain", StormData$EVTYPE,ignore.case = TRUE)] <- "Rain"
StormData$EVTYPE[grepl("freez|cold|snow|chill|winter|blizzard|sleet|glaze|frost|hypothermia|wintry*|low temp*|mixed precip|ice", 
                      StormData$EVTYPE,ignore.case = TRUE)] <- "Cold Weather"
StormData$EVTYPE[grepl("TROPICAL.Storm|typhoon|*depression", StormData$EVTYPE,ignore.case = TRUE)] <- "Tropical Storm"
StormData$EVTYPE[grepl("(hurri|opal)", StormData$EVTYPE,ignore.case = TRUE)] <- "Tropical Storm"
StormData$EVTYPE[grepl("RIP CURREN(T|TS)", StormData$EVTYPE,ignore.case = TRUE)] <- "Rip Current"
StormData$EVTYPE[grepl("*Current|*SURF|*SEAS|marine*|*swells|*waves|waterspout|*tide", StormData$EVTYPE,ignore.case = TRUE)] <- "Nautical"
StormData$EVTYPE[grepl("heat", StormData$EVTYPE,ignore.case = TRUE)] <- "Heat"
StormData$EVTYPE[grepl("Storm SURGE", StormData$EVTYPE,ignore.case = TRUE)] <- "Storm Surge"
```

Now we can take a look at the Property Damage and Crop Damage Exponents to get a sense of they are used in the data set. We start by examining and then transforming the Property Damage Exponents variable.

``` {r Data Cleaning 2}
## Clean the Property and Crop Damage Data
unique(StormData$PROPDMGEXP)

## Convert exponents into numeric values
StormData$PROPDMG[StormData$PROPDMGEXP == "K"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "K"] * 1000
StormData$PROPDMG[StormData$PROPDMGEXP == "M"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "M"] * (10^6)
StormData$PROPDMG[StormData$PROPDMGEXP == "H"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "H"] * 100
StormData$PROPDMG[StormData$PROPDMGEXP == "h"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "h"] * 100
StormData$PROPDMG[StormData$PROPDMGEXP == ""] <- StormData$PROPDMG[StormData$PROPDMGEXP == ""] * 1
StormData$PROPDMG[StormData$PROPDMGEXP == "B"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "B"] * (10^9)
StormData$PROPDMG[StormData$PROPDMGEXP == "m"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "m"] * (10^6)
StormData$PROPDMG[StormData$PROPDMGEXP == "0"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "0"] * 1
StormData$PROPDMG[StormData$PROPDMGEXP == "1"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "1"] * 10
StormData$PROPDMG[StormData$PROPDMGEXP == "2"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "2"] * 100
StormData$PROPDMG[StormData$PROPDMGEXP == "3"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "3"] * 1000
StormData$PROPDMG[StormData$PROPDMGEXP == "4"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "4"] * (10^4)
StormData$PROPDMG[StormData$PROPDMGEXP == "5"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "5"] * (10^5)
StormData$PROPDMG[StormData$PROPDMGEXP == "6"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "6"] * (10^6)
StormData$PROPDMG[StormData$PROPDMGEXP == "7"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "7"] * (10^7)
StormData$PROPDMG[StormData$PROPDMGEXP == "8"] <- StormData$PROPDMG[StormData$PROPDMGEXP == "8"] * (10^8)
StormData$PROPDMG[StormData$PROPDMGEXP == "+"] <- 0
StormData$PROPDMG[StormData$PROPDMGEXP == "-"] <- 0
StormData$PROPDMG[StormData$PROPDMGEXP == "?"] <- 0
head(StormData[,c("EVTYPE","PROPDMG","PROPDMGEXP")])
```

As you can see, NOAA's meterologists used a variety of means to indicate exponents including "K" and "3" for thousands, and "B" and "8" for billions. We can now perform the same tranformation for the Crop Damage Exponent.

``` {r Data Cleaning 3}
StormData$CROPDMG[StormData$CROPDMGEXP == "M"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "M"] * (10^6)
StormData$CROPDMG[StormData$CROPDMGEXP == "K"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "K"] * 1000
StormData$CROPDMG[StormData$CROPDMGEXP == "m"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "m"] * (10^6)
StormData$CROPDMG[StormData$CROPDMGEXP == "B"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "B"] * (10^9)
StormData$CROPDMG[StormData$CROPDMGEXP == "k"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "k"] * 1000
StormData$CROPDMG[StormData$CROPDMGEXP == "0"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "0"] * 1
StormData$CROPDMG[StormData$CROPDMGEXP == "2"] <- StormData$CROPDMG[StormData$CROPDMGEXP == "2"] * 100
StormData$CROPDMG[StormData$CROPDMGEXP == ""] <- StormData$CROPDMG[StormData$CROPDMGEXP == ""] * 1
StormData$CROPDMG[StormData$CROPDMGEXP == "?"] <- 0
head(StormData[,c("EVTYPE","CROPDMG","CROPDMGEXP")])
```

We're now in a position to take a look at the data and answer two questions:  
1 - Across the US, which event typs (EVTYPE variable) are most harmful to population health?  
2 - Across the US, which event types have the greatest economic conseqences?

## Section Three: Data Analysis and Results

